Pleiotropy of gene regulatory variants increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2022-11-13 ms/ms03_results.tex
MS

To better understand the role of gene regulatory variants for variant pleiotropy, I developed a pipeline to systematically colocalize eQTLs and GWAS variants (Section Methods).

127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many tissues and cell types \citep{2021.Alasoo.Kerimov}.

Public GWAS summary statistics were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS studies with at least 10000 subjects, 2000 controls and 2000 cases (Supplementary table 2).
%
These filters resulted in 418 GWAS (Supplementary table 2).
%
I developed a colocalization pipeline that explored all the 51 625 combinations of eQTL studies and GWAS (Section Methods).
%
This analysis tested 139,069 unique variants and ? potential variant/GWAS/eQTL sample and e-gene colocalizations from 418 GWAS studies (OSF URL).
%select count(*) from (select DISTINCT chrom,pos,rsid FROM coloc); 139,069
%select count(*) from (select DISTINCT chrom,pos,rsid FROM coloc where `PP.H4.abf`>=0.8); 103,551
%select count(*) from (select DISTINCT chrom,pos,rsid,gwas_id,eqtl_id,egene FROM coloc);
%select count(*) from (select DISTINCT chrom,pos,rsid,gwas_id,eqtl_id,egene FROM coloc where `PP.H4.abf`>=0.8); 2,477,640
%
I found 103,551 variants or from ? GWAS that colocalized with at least one eQTL.
%select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.8);
%
%TODO add PMID to supplementary table with loci_explained_perc
For GWAS studies with more than 10 loci, the explained of loci variate from 0\% to 77\% with an average of 33\% (Supplementary table ST6).
%
Autoimmune diseases such as inflammatory bowel disease, Crohn's disease and Ulcerative colitis showed a colocalization average of 45.4\% in agreement with previous work \citep{2021.Li.Mu.GenomeBiology.impactcelltype}.
%
Selection of colocalizations with PP.H4.abf greater than 0.8 resulted in 9 758 variants and 143 119 colocalizations (Supplementary table 3).

@@

%
In the EBI eQTL Catalogue, immune cell types are annotated to great detail while other tissues are annotated at lower resolution.
%
Therefore I manually defined 36 categories to annotate eQTL biological samples at equivalent resolution (Supplementary table 1).

These category annotations were used to investigate the variant pleiotropy for GWAS traits.
%
Colocalized eQTL/GWAS variants were classified according to the number of GWAS categories they belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
Pleiotropic variants aggregate in some cytobands such as 3q23, 5q31.1, 11q13.5, 12q24.12 and 15q26.1 (Table \ref{tab:pleitropic_variants} and Supplementary table 3).

>>>>

To better understand the role of gene regulatory variants for variant pleiotropy, I developed a pipeline to systematically colocalize eQTLs and GWAS variants (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many tissues and cell types \citep{2021.Alasoo.Kerimov}.

Public GWAS summary statistics were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS studies with at least 10,000 subjects, 2,000 controls and 2,000 cases (Supplementary table 2).
%
These filters resulted in 335 GWAS with 10,627 clumped leading variants and a p-value below 5e-8 (Supplementary table 2).
% sqlite
% select count(distinct gwas_id) from tophits; 335
% select count(*) from (select DISTINCT chrom, pos, rsid, nea, ea from tophits) limit 50; 10627

I found eQTL colocalizations for 3,849 or 36$\%$ of the leading GWAS variants with 2,348 eQTL genes and 127 eQTL samples.
% sqlite
% select count(distinct tophits.rsid) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8; 3849
% alternative: select count(distinct tophits.rsid) from tophits, coloc where coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8; 3849
% select count(distinct coloc.egene) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8
% select count(distinct coloc.eqtl_id) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8

The largests classes of diseases by number of leading GWAS variants are autoimmune diseases (2107), cancer of breast (1976) and allergy (1321)
%
and the percentage of variants with a colocalized eQTL is 37\%, 21\% and 42\%, respectively (Supplementary table ST6).
%
%/* count tophits.rsid group by gwas_annot.gwas_class where coloc.`PP.H4.abf`>=0.8*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8 GROUP by gwas_annot.gwas_class;
%/* count tophits.rsid group by gwas_annot.gwas_class*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id GROUP by gwas_annot.gwas_class order by gwas_annot.gwas_class;
%
For the autoimmune diseases, this is in agreement with previous work \citep{2021.Li.Mu.GenomeBiology.impactcelltype}.

For the autoimmune diseases, ? of the ? leading variants had an annotation with probability PP.H4.abf$\ge$0.8.
% sqlite
% select count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_class='Autoimmune dis.' and gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8;
% 778
% grouped H4 0.8: select count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8 GROUP by  gwas_annot.gwas_class;
%
This is perc of the
% grouped all: select count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid GROUP by  gwas_annot.gwas_class;%
%
%%TODO add PMID to supplementary table with loci_explained_perc
%For GWAS studies with more than 10 loci, the explained of loci variate from 0\% to 77\% with an average of 33\% (Supplementary table ST6).
%%
%Autoimmune diseases such as inflammatory bowel disease, Crohn's disease and Ulcerative colitis showed a colocalization average of 45.4\% in agreement with previous work \citep{2021.Li.Mu.GenomeBiology.impactcelltype}.
%%
%Selection of colocalizations with PP.H4.abf greater than 0.8 resulted in 9 758 variants and 143 119 colocalizations (Supplementary table 3).

This pipeline has resulted in 103,551 variants with or from 246 GWAS that colocalized (PP.H4.abf 0.8) with at least one eQTL.
%select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.8);
%
This is a useful annotation ressource that I have made available through a web application (TODO).
%
%This analysis tested 139,069 unique variants in windows of one million nucleotides around the GWAS leading variants and ? potential variant/GWAS/eQTL sample and e-gene colocalizations from 418 GWAS studies (OSF URL).
%select count(*) from (select DISTINCT chrom,pos,rsid FROM coloc); 139,069
%select count(*) from (select DISTINCT chrom,pos,rsid,gwas_id,eqtl_id,egene FROM coloc); 17,595,868
%
%There are 103,551 variants from 246 GWAS with a least one colocalized eQTL and probability PP.H4.abf$\ge$0.8.
%select count(DISTINCT gwas_id) from coloc where `PP.H4.abf`>=0.8; 246
%select count(*) from (select DISTINCT chrom,pos,rsid FROM coloc where `PP.H4.abf`>=0.8); 103,551
%
%select count(*) from (select DISTINCT chrom,pos,rsid,gwas_id,eqtl_id,egene FROM coloc where `PP.H4.abf`>=0.8); 2,477,640

@@

In the EBI eQTL Catalogue, immune cell types or neural tissues are annotated to great detail while other tissues are annotated at lower resolution.
%
Therefore I manually defined 36 categories to annotate eQTL samples at equivalent resolution (Supplementary table 1).

These category annotations were used to investigate the variant pleiotropy for GWAS traits.
%
Colocalized eQTL/GWAS variants were classified according to the number of GWAS categories they belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
%
These aggregated variants were highly pleiotropic.
%
For instance the variant rs2522051, which is located downstream of the IRF1 gene,

re located upstream and downstream of the IRF1 gene, respectively.

%Pleiotropic variants aggregate in some cytobands such as 3q23, 5q31.1, 11q13.5, 12q24.12 and 15q26.1 (Table \ref{tab:pleitropic_variants} and Supplementary table 3).
Pleiotropic variants aggregate in some cytobands such as 5q31.1, 19q13.33, 3q23, 9p21.3 and 15q24.1 (Table \ref{tab:pleitropic_variants} and Supplementary table 3).

% 5q31.1
